Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2021-09-19 publications/iaai21/dora_outlier_detection_iaai.tex
Reorder table 1 by size. #42
Update random results for astro and fmnist. #42
Trim length by making text more concise. #42

\begin{abstract}
Automatic detection of out-of-distribution (OOD) samples or features is
universally needed when working with scientific datasets. For example,
OOD detection methods can be used for cleaning datasets or flagging
novel samples in order to guide instrument acquisition or scientific analysis.
We present Domain-agnostic Outlier Ranking Algorithms (DORA) to provide
a configurable pipeline that makes it easy for scientists to apply and evaluate
outlier detection methods in a variety of domains. DORA allows users to 
configure outlier detection experiments by specifying the location of their 
dataset(s), the input data type, feature extraction methods, and which 
algorithms should be applied. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
 (DEMUD), principal component analysis (PCA), Reed-Xiaoli (RX) detector, 
 Local RX, negative sampling, and probabilistic autoencoder. Given a set of
  data samples, each algorithm assigns an outlier score to each sample. DORA
   provides results organization and visualization modules to help users 
   process the results, including sorting samples by outlier score, evaluating 
   outlier recall for a set of known/labeled outliers, clustering groups of 
   similar outliers together, and causal graphs. To demonstrate the 
   cross-domain applicability of DORA, we conducted outlier detection
    experiments for three real-world use cases: 1) cleaning noisy ground-truth 
    datasets (Earth Science), 
     2) flagging non-astrophysical objects in the Dark Energy Survey 
     (astrophysics), and 3) novelty-guided onboard targeting of instruments
      for Mars rovers (planetary science). [We showed that ...]
\end{abstract}

@@

\section{Related Work}
Methods for outlier detection have been 
surveyed extensively in prior work
~\cite{pimentel2014review,chandola2009anomaly,markou2003novelty,
markou2003novelty2}.
Methods can be differentiated primarily based on how they define outlierness.
%
Reconstruction-based methods aim to construct a model of a dataset by learning 
a mapping between the data samples and a lower-dimensional representation that
 minimizes the loss between the input and its reconstruction from the 
 low-dimensional representation~\citep{kerner2020comparison}. The 
 reconstruction error is used to compute the outlier score because samples
  that are unlike the data used to fit the model will be more poorly 
  reconstructed compared to inliers. Reconstruction-based methods
  include PCA~\cite{jablonski2015principal}, 
  autoencoders~\cite{richter2017safe,park2018multimodal}, and
  generative adversarial networks~\cite{akcay2018ganomaly}.
%
 Distance-based methods score outliers based on their distance from some
``background'' which can be defined in a variety of ways. For example,
the Reed-Xiaoli (RX) detector computes the 
Mahalanobis distance between each sample and the background dataset
defined by its mean and covariance matrix~\cite{reed1990adaptive}.
%
Sparsity-based methods, such as the isolation forest~\cite{liu2008isolation}
and local outlier factor\cite{breunig2000lof},
 score outliers based on how isolated or sparse they
are in a given feature space. 
%
Probability distribution and density based methods estimate the underlying 
distribution or probability density of a dataset and score samples based on
their likelihood under the learned distribution. Examples include the probabilistic
autoencoder, which scores samples based on the log likelihood under the learned
latent space distribution~\cite{bohm2020probabilistic}, Gaussian mixture 
Models, and kernel density estimators~\cite{chandola2009anomaly}. 
%
Other methods treat detection of outliers as a supervised classification 
problem, usually with only one class constituted by the inliers or 
known normal samples.
Such methods include one-class support vector 
machines~\cite{scholkopf1999support},
support vector data description~\cite{pimentel2014review},
and negative sampling~\cite{sipple:neg-sampling20}.

@@

\paragraph{Data types.} We chose to initially implement data loaders for four
data types that are commonly used by the machine learning and domain science
communities: time series, feature vectors, images (grayscale or RGB), and 
$N$-band rasters. $N$-band rasters are essentially images in which every pixel
is associated with a location defined by a geographic coordinate reference 
system (e.g., latitude/longitude in degrees). Most satellite data are 
distributed as rasters; common formats include GeoTIFF, NetCDF, and 
HDF. A data loader for each data type locates the data by the path(s)
defined in the configuration file and loads samples into a dictionary of numpy
 arrays indexed by the sample id. This \texttt{data\_dict} is then passed to each of the ranking algorithms specified in 
 the configuration file.

\paragraph{Outlier ranking algorithms.} We implemented 7 algorithms for 
scoring and 
ranking samples by outlierness. All algorithms are unsupervised.
We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may have 
better performance for different datasets and use cases. 
%We describe each
%approach to scoring outliers and the associated methods below.

@@

We constructed three datasets to evaluate the utility of DORA and the
performance of each ranking algorithm for a variety of scientific domains
 (astrophysics, planetary science, and Earth science). We also included a 
 benchmark dataset that uses MNIST and Fashion-MNIST. Table \ref{tab:datasets}
 summarizes the number of unlabeled samples used for training and labeled
 samples used for evaluation for each dataset. We describe each in detail
 below.

\subsubsection{Astrophysics: objects in Dark Energy Survey.}

Astronomical data sets are large and growing. Large modern optical
imaging surveys are producing catalogs for of order $10^8$ stars
and galaxies, with dozens or hundreds of distinct measured features
for each entry. Discovery science becomes difficult at this data
volume: the scale is too large for expert human inspection, and separating real
astrophysical anomalies from non-astrophysical sources like detector artifacts or
satellite trails is a challenging problem for current
methods.

The Dark Energy Survey (DES), an ongoing imaging survey of
$5000~{\rm deg}^2$ of the southern sky conducted from the Cerro-Tololo
Inter-American Observatory in Chile \cite{DESY1}. The galaxy catalogs produced from
DES imaging have provided some of the strongest constraints to date on
the physical properties of dark energy and the accelerated expansion
of the Universe. The first version of this catalog, released in June 2018, incorporated
only cuts on signal-to-noise, resolution, masks against known detector
anomalies and data quality indicators, and the automated data quality
flags produced during processing. In December 2019, the full catalog
was released after 18 months of extensive manual vetting. Therefore,
we were able to use the second version of the catalog as a validation
set for anomaly detection on the first version.

We compared all methods on a data set that consists of 100K galaxy objects
observed by the Dark Energy Survey (DES) sampled from the initial June 2018 release.
25,339 objects from this 100K set that did not appear in the later Dec 2019
release were likely eliminated during the manual vetting process and were labeled as outliers.
Note, that the remaining 74,661 objects may also contain outliers, but we assume them to be
inliers in this experiment.

We used publicly-available photometry from the $g-$, $r-$, $i-$ and $z-$band DES exposures.  We
transformed the photometry into luptitudes\footnote{A
  ``Luptitude''~\cite{lupton99} is an arcsinh-scaled flux, with
  properties quantitatively equal to
  traditional astronomical magnitudes for bright sources, but which
  gracefully handles non-detections and negative fluxes.}.  The inputs features used
  were the $r$-band luptitude, colors computed as band differences between
$g-r$, $i-r$, and $z-r$, and associated observational errors, for a total of eight features.
%

@@

We simulated the operational setting in which the rover has observed
targets up through mission day (sol) $s$ and the goal is to rank all
subsequent targets (after sol $s$) to inform which recent targets
merit further study.  Our rover image data covers sols \num{1343}
to \num{1703}.  We partitioned the images chronologically to assess
outlier detection in the 10 most recent sols, using ``prior'' set
$D_{1343-1693}$ ($n=992$) and ``assessment'' set $D_{1694-1703}$
($n=58)$ for evaluation.

We collaborated with an MSL science planning team member to
independently review the targets in $D_{1694-1703}$ and identify those
that were 
considered novel by the mission ($n_{outlier} = 9$).  Our goal for
this application is to assess how well the selections made by each
algorithm correlate with human novelty judgments to determine which
methods would be most suitable for informing onboard decisions about
follow-up observations.

\subsubsection{Earth: satellite time series for ground-truth observations.}
Many Earth science applications using satellite Earth observation (EO) data
 require ground-truth observations for understanding and modeling ground-
 identified objects in the satellite observations. These ground observations
 also serve as labels that are paired with satellite data inputs for machine 
 learning models. For example, a model trained to classify crop types in
 satellite observations requires geo-referenced labels of crop type annotated
 in the field~\cite{tseng2021learning}. A widespread challenge of using ground-
 annotated datasets is that there are often points with erroneous location or 
 label information (e.g., due to GPS location error or human entry error) that
 need to be cleaned before the dataset can be used for machine learning or
 other downstream use cases. Automatically detecting these outliers could save 
 substantial time required for cleaning datasets and improve the performance of
 downstream analyses that rely on high-quality datasets. 
 %
 We used a dataset of ground
 observations of fall army worm (FAW) in maize crops collected by the UN Food
 and Agriculture Organization (FAO). This dataset includes 6,757 samples
 with location (latitude/longitude) metadata primarily in Africa and Southeast
 Asia. Most locations coincide with crop
 fields but there are many outliers that coincide with other land cover types 
 such as water, buildings, or forests. We constructed an evaluation set 
 consisting of all samples in Kenya ($n=113$) and manually annotated whether 
 each sample was considered an inlier (i.e., located in a crop field) or not 
 using high-resolution Planet satellite images in Collect Earth 
 Online~\cite{planetdata} ($n_{inlier}=76$, $n_{outlier}=37$). We used Google
 Earth Engine to export the Sentinel-1 synthetic aperture radar (SAR) monthly
 time series for each sample location, using data from year each sample was 
 collected. We used SAR data because it is sensitive to ground texture and
  penetrates clouds, which is important for the often-cloudy region covered
  by the dataset. Our goal for this application is to assess how well the 
  selections made by each algorithm correlate with outliers determined by
  visual inspection of the satellite images.
% descoped for lack of time and space
% \todo{If we have time, I want to also evaluate the downstream classifier
% performance before and after removing top N outliers with DORA.}

@@

\subsubsection{Benchmark: MNIST and Fashion-MNIST.}
We included MNIST and F-MNIST to demonstrate the capabilities of DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a combined set of 2,000 images from MNIST and 
F-MNIST as the evaluation set. Well-discriminating algorithms should assign 
higher outlier scores to samples from MNIST than F-MNIST.

\begin{table}
  \caption{Number of samples in the training and test sets for each dataset.}
  \label{tab:datasets}
  \centering
  \begin{tabular}{l|c|cc}
    \hline
    Dataset & Training & \multicolumn{2}{c}{Test}\\
     & Unlabeled &  Outliers &  Inliers \\

    \hline
    Astrophysics & 100,000   &  25,339 &  ---- \\
    Earth & 6,757 & 37 & 76  \\
    Planetary & 992 &  9 & 49 \\
    F-MNIST & 60,000 & 1,000 & 1,000  \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications and F-MNIST
   benchmark; the best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|cccc}
    \hline
    Algorithm & Astro. & Earth & Planetary & F-MNIST \\
    \hline
    PCA               & \textbf{0.42} & 0.41 & 0.44 & \textbf{0.84} \\
    DEMUD             & ---- & 0.41 & 0.44 & \textbf{0.84} \\
    \hline
    RX                & 0.40 & 0.43 & ---- & 0.82 \\
    LRX               & ----& ---- & 0.33 & 0.56 \\
    \hline
    Iso. Forest  & 0.34 & 0.46 & \textbf{0.56} & 0.74 \\
    \hline
    PAE               & 0.35 & 0.41 & 0.44 & 0.83 \\
    Neg. Sampling & 0.32 & \textbf{0.49} & 0.33 & 0.43 \\
    \hline
    Random            & 0.26 & 0.32 & 0.14 & 0.49 \\
    \hline
  \end{tabular}
\end{table}

>>>>

\begin{abstract}
Automatic detection of out-of-distribution (OOD) samples or features is
universally needed when working with scientific datasets. For example,
OOD detection methods can be used for cleaning datasets or flagging
novel samples in order to guide instrument acquisition or scientific analysis.
We present Domain-agnostic Outlier Ranking Algorithms (DORA) to provide
a configurable pipeline that makes it easy for scientists to apply and evaluate
outlier detection methods in a variety of domains. DORA allows users to 
configure outlier detection experiments by specifying the location of their 
dataset(s), the input data type, feature extraction methods, and which 
algorithms should be applied. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
 (DEMUD), principal component analysis (PCA), Reed-Xiaoli (RX) detector, 
 Local RX, negative sampling, and probabilistic autoencoder. Given a set of
  data samples, each algorithm assigns an outlier score to each sample. DORA
   provides results organization and visualization modules to help users 
   process the results, including sorting samples by outlier score, evaluating 
   outlier recall for a set of known/labeled outliers, clustering groups of 
   similar outliers together, and causal graphs. To demonstrate the 
   cross-domain applicability of DORA, we conducted outlier detection
    experiments for three real-world use cases: 1) cleaning noisy ground-truth 
    datasets (Earth Science), 
     2) flagging non-astrophysical objects in the Dark Energy Survey 
     (astrophysics), and 3) novelty-guided onboard targeting of instruments
      for Mars rovers (planetary science). \todo{[We showed that ...]}
\end{abstract}

@@

\section{Related Work}
Methods for outlier detection have been 
surveyed extensively and can be differentiated primarily based on how they 
score outliers
~\cite{pimentel2014review,chandola2009anomaly,markou2003novelty,
markou2003novelty2}.
%
Reconstruction-based methods construct a model of a dataset by learning 
a mapping between the input data and a lower-dimensional representation that
 minimizes the loss between the input and its reconstruction from the 
 low-dimensional representation~\citep{kerner2020comparison}. The 
 reconstruction error is used as the outlier score because samples
  that are unlike the data used to fit the model will be more poorly 
  reconstructed compared to inliers. Reconstruction-based methods
  include PCA~\cite{jablonski2015principal}, 
  autoencoders~\cite{richter2017safe,park2018multimodal}, and
  generative adversarial networks~\cite{akcay2018ganomaly}.
%
 Distance-based methods score outliers based on their distance from a
``background'' which can be defined in a variety of ways. For example,
the Reed-Xiaoli (RX) detector computes the 
Mahalanobis distance between each sample and the background dataset
defined by its mean and covariance matrix~\cite{reed1990adaptive}.
%
Sparsity-based methods such as isolation forest~\cite{liu2008isolation}
and local outlier factor~\cite{breunig2000lof}
 score outliers based on how isolated or sparse samples 
are in a given feature space. 
%
Probability distribution and density based methods estimate the underlying 
distribution or probability density of a dataset and score samples based on
their likelihood. Examples include the probabilistic
autoencoder, which scores samples based on the log likelihood under the learned
latent space distribution~\cite{bohm2020probabilistic}, Gaussian mixture 
Models, and kernel density estimators~\cite{chandola2009anomaly}. 
%
Other methods formulate outlier detection as supervised classification, 
usually with only one class constituted by the inliers or 
known normal samples.
Such methods include one-class support vector 
machines~\cite{scholkopf1999support},
support vector data description~\cite{pimentel2014review},
and negative sampling~\cite{sipple:neg-sampling20}.

@@

\paragraph{Data types.} We chose to implement data loaders for four
data types that are commonly used by the machine learning and domain science
communities: time series, feature vectors, images (grayscale or RGB), and 
$N$-band rasters. $N$-band rasters are images or grids in which every pixel
is associated with a location (e.g., latitude/longitude in degrees); 
most satellite data are distributed as rasters.
A data loader for each data type locates the data by the path(s)
defined in the configuration file and loads samples into a dictionary of numpy
 arrays indexed by the sample id. This \texttt{data\_dict} is then passed to each of the ranking algorithms.

\paragraph{Outlier ranking algorithms.} We implemented 7 unsupervised 
algorithms for scoring and 
ranking samples by outlierness. 
We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may  
perform better for different datasets and use cases. 
%We describe each
%approach to scoring outliers and the associated methods below.

@@

We constructed three datasets to evaluate the utility of DORA and 
algorithm performance for a variety of scientific domains
 (astrophysics, planetary science, and Earth science). We also included a 
 benchmark dataset that uses MNIST and Fashion-MNIST. Table \ref{tab:datasets}
 summarizes the number of unlabeled samples used for training evaluation for
 each dataset. We describe each dataset in detail below.

\subsubsection{Astrophysics: objects in Dark Energy Survey.}
Astronomical data sets are large and growing. Large modern optical
imaging surveys are producing catalogs of order $10^8$ stars
and galaxies, with dozens or hundreds of distinct measured features
for each entry. Discovery science becomes difficult at this data
volume: the scale is too large for expert human inspection, and separating real
astrophysical anomalies from non-astrophysical sources like detector artifacts or
satellite trails is a challenging problem for current methods.

The Dark Energy Survey (DES) is an ongoing imaging survey of
$5000~{\rm deg}^2$ of the southern sky from the Cerro-Tololo
Inter-American Observatory in Chile \cite{DESY1}. The resulting galaxy catalogs produced have provided some of the strongest constraints to date on
the physical properties of dark energy and accelerated expansion
of the Universe. The first version of this catalog, released June 2018, incorporated
only cuts on signal-to-noise and resolution, masks against known detector
anomalies and data quality indicators, and the automated data quality
flags produced during processing to filter outliers. 
In December 2019, the full catalog
was released after 18 months of extensive manual vetting. We used the samples 
that were removed in the second version of the catalog as a set of known
outliers for evaluating anomaly detection methods on the first version.

We compared all methods on a dataset of 100K galaxy objects
observed by the Dark Energy Survey (DES) sampled from the initial June 2018 release.
We labeled the 25,339 objects from this 100K set that did not appear in the later December 2019
release, thus were likely eliminated during the manual vetting process, 
as outliers.
While the remaining 74,661 objects may also contain outliers, we assume them to be inliers in this experiment. We used publicly-available photometry from the $g-$, $r-$, $i-$ and $z-$band DES exposures.  We
transformed the photometry into luptitudes\footnote{A
  ``Luptitude''~\cite{lupton99} is an arcsinh-scaled flux, with
  properties quantitatively equal to
  traditional astronomical magnitudes for bright sources, but which
  gracefully handles non-detections and negative fluxes.}.  The input features 
  were the $r$-band luptitude, colors computed as band differences between
$g-r$, $i-r$, and $z-r$, and associated observational errors, for a total of eight features.
%

@@

We simulated the operational setting in which the rover has observed
targets up through mission day (sol) $s$ and the goal is to rank all
subsequent targets (after sol $s$) to inform which recent targets
merit further study.  Our rover image data covers sols \num{1343}
to \num{1703}.  We partitioned the images chronologically to assess
outlier detection in the 10 most recent sols, using ``prior'' set
$D_{1343-1693}$ ($n=992$) and ``assessment'' set $D_{1694-1703}$
($n=58)$ for evaluation. We collaborated with an MSL science team member to
independently review the targets in $D_{1694-1703}$ and identify those
considered novel by the mission ($n_{outlier} = 9$).  Our goal for
this application is to assess how well the selections made by each
algorithm correlate with human novelty judgments to determine which
methods would be most suitable for informing onboard decisions about
follow-up observations.

\subsubsection{Earth: satellite time series for ground observations.}
Many Earth science applications using satellite Earth observation (EO) data
 require ground-truth observations for identifying and modeling ground-
 identified objects in the satellite observations. These ground observations
 also serve as labels that are paired with satellite data inputs for machine 
 learning models. For example, a model trained to classify crop types in
 satellite observations requires ground-annotated labels of crop type~\cite{tseng2021learning}. A widespread challenge for ground-
 annotated labels is that there are often points with erroneous location or 
 label information (e.g., due to GPS location error or human entry error) that
 need to be cleaned before the labels can be used for machine learning or
 other downstream uses. Automatically detecting these outliers could save 
 substantial time required for cleaning datasets and improve the performance of
 downstream analyses that rely on high-quality datasets.

 We used a dataset of ground annotations of maize crops collected by the UN Food
 and Agriculture Organization (FAO). This dataset includes 6,757 samples
 with location (latitude/longitude) metadata primarily in Africa and Southeast
 Asia. Most locations coincide with crop
 fields but there are many outliers that coincide with other land cover types 
 such as water, buildings, or forests. We constructed an evaluation set 
 of all samples in Kenya ($n=113$) and manually annotated whether 
 each sample was located in a crop field (inlier) or not (outlier)
 using high-resolution satellite images in Collect Earth 
 Online ($n_{inlier}=76$, $n_{outlier}=37$). We used Google
 Earth Engine to export the Sentinel-1 synthetic aperture radar (SAR) monthly
 time series for each sample location, using data from the year each sample was 
 collected. We used SAR data because it is sensitive to ground texture and
  penetrates clouds, which is important for the often-cloudy region covered
  by the dataset. Our goal for this application was to assess how well the 
  selections made by each algorithm correlate with outliers determined by
  visual inspection of the satellite images.
% descoped for lack of time and space
% \todo{If we have time, I want to also evaluate the downstream classifier
% performance before and after removing top N outliers with DORA.}

@@

\subsubsection{Benchmark: MNIST and Fashion-MNIST.}
We used MNIST and F-MNIST to demonstrate the capabilities of DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images from MNIST (outliers)
 and 1,000 from F-MNIST (inliers). 
 %Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.

\begin{table}
  \caption{Number of samples in the training and test sets for each dataset,
  ordered by dataset size.}
  \label{tab:datasets}
  \centering
  \begin{tabular}{l|c|cc}
    \hline
    Dataset & Training & \multicolumn{2}{c}{Test}\\
     & Unlabeled &  Outliers &  Inliers \\

    \hline
    Planetary & 992 &  9 & 49 \\
    Earth & 6,757 & 37 & 76  \\
    F-MNIST & 60,000 & 1,000 & 1,000  \\
    Astrophysics & 100,000   &  25,339 &  74,661 \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications and F-MNIST
   benchmark; the best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|cccc}
    \hline
    Algorithm & Astro. & Earth & Planetary & F-MNIST \\
    \hline
    PCA               & \textbf{0.42} & 0.41 & 0.44 & \textbf{0.84} \\
    DEMUD             & ---- & 0.41 & 0.44 & \textbf{0.84} \\
    \hline
    RX                & 0.40 & 0.43 & ---- & 0.82 \\
    LRX               & ----& ---- & 0.33 & 0.56 \\
    \hline
    Iso. Forest  & 0.34 & 0.46 & \textbf{0.56} & 0.74 \\
    \hline
    PAE               & 0.35 & 0.41 & 0.44 & 0.83 \\
    Neg. Sampling & 0.32 & \textbf{0.49} & 0.33 & 0.43 \\
    \hline
    Random            & 0.25 & 0.32 & 0.14 & 0.50 \\
    \hline
  \end{tabular}
\end{table}
